BlueTracer: \\ a Robust API Tracer for Evasive Malware

### 2018-07-02 implementation.tex
Implementation Section Done

In light of what just stated, BlueTracer's integration with BluePill is straightforward, since it essentially complements the \textit{Target platform simulation layer}, as it can be observed from \textit{Figure 3.1}.

@@

In Pin, the Native API is intercepted right before \texttt{sysenter} is executed. This could be inferred by the fact that, in \texttt{TraceSysCallEntry}, the instruction pointer (obtainable via the \texttt{PIN\_GetContextReg} API) contains \texttt{sysenter}'s address. In light of this, in addition to the fact that each Native API calls \texttt{KiFastSystemCall} without setting a stack frame, the application's stack during the execution of \texttt{TraceSysCallEntry} is in the following state (\textit{Figure 3.4}).
\\
\begin{figure}[h]
\centering
\includegraphics[width=0.7\textwidth]{Figures/StackMainImage.pdf}
\caption{\textit{Stack before} \texttt{sysenter} \textit{execution}}
\end{figure}

Having outlined the general situation, it is finally possible to describe how the main image check is carried out:

@@

		// Array of arguments
		sysinfo_arg_t arg[MAX_ARGS_IN_ENTRY];

		...
	} syscall_info_t;
\end{lstlisting}

@@

Even for \texttt{syscall\_tracer}, each field is self-explanatory. The only one which was not encountered before is the \texttt{syscall\_counter} field. The rationale behind this field is that, in the log, each traced call has a unique integer associated to it, utilized to group together the information relative to the call in post-processing. The next identifier to be assigned is stored in the \texttt{call\_number} field of the thread local storage (\textit{Listing 3.1}). Therefore, \texttt{syscall\_counter} simply contains the unique identifier associated to the Native API in the log. The arguments of a specific Native API are represented via the \texttt{drsys\_arg\_t} struct (\textit{Listing 3.6}).
\\
\begin{lstlisting}[caption={\texttt{struct} representing the an argument of the Native API being traced},captionpos=b]
typedef struct _drsys_arg_t {

@@

\texttt{drsys\_arg\_t}'s fields are mostly the same of \texttt{sysinfo\_arg\_t}, with three additions:
\begin{itemize}
\item \texttt{pre}, which is a boolean flag set to \texttt{true} when the logging is occurring before the Native API is executed (i.e. in \texttt{TraceSysCallEntry} and set to false when the logging is occurring after the Native API is executed (i.e. in \texttt{TraceSysCallExit}).
\item \texttt{enum\_name}, a string containing the symbolic name of the enumeration a named constant belongs to. In most cases, this is the symbolic name associated to the first constant of the enumeration. Such field was introduced to decouple type name and enumeration name.
\item \texttt{value}, an \texttt{ADDRINT}\footnote{The ADDRINT type is defined by Pin and represents a memory address} containing the value of the argument. This is obtained through the \texttt{PIN\_GetSyscallArgument} API, which takes as input the context before the execution of the system call, the system call standard and the ordinal number of the argument whose value is requested.  
\end{itemize}

@@

\item Finally, if the argument's type is not a primitive one, it is checked if it is \texttt{null} and if it is a complex output parameter being traced before the Native API execution, situation in which only the address stored in the pointer is logged. If none of these two scenarios are true, then the tracer finds itself in the situation where it is required to log the value of a complex type. This is performed through the invocation of \texttt{print\_known\_compound}. As of right now, such function differentiates between four complex types, namely \texttt{UNICODE\_STRING}, \texttt{OBJECT\_ATTRIBUTES}, \texttt{IO\_STATUS\_BLOCK} and \texttt{LARGE\_INTEGER}, and logs the argument's value accordingly. If the argument's type is a complex type that is not supported, the symbolic value \texttt{<NYI>} (not yet implemented) is recorded.

@@

To conclude this section, let us show how a traced Native API looks like in the log file (\textit{Figure 3.5}). As it can be observed, the adopted log format is quite intuitive. In fact, the idea was to have a log which is both easy to read by the analyst but also easy to parse and post-process.      
\\

@@

In Microsoft Windows, APIs are implemented as functions provided by a set of dynamic-link libraries. These, also known as DLLs, constitute Microsoft's way of implementing the concept of shared libraries in the Windows operating system. Therefore, the functions they make available can be employed by different applications \cite{DLL}. Windows APIs offer a plethora of functionalities which belong to many different categories, ranging from base services (e.g. file management) to user interface functions and network operations \cite{WinAPI}. In light of this, a Windows API has rich semantic information associated to it and tracing the sequence of APIs a malware calls provides analysts with a very high-level view of the sample's behavior. This is why BlueTracer's API tracer component is undoubtedly the most important one. 
\\
\begin{figure}[h]
\centering
\includegraphics[scale=0.662]{Figures/APITracing.pdf}
\caption{\textit{APIs Tracing Workflow}}
\end{figure}

When tracing APIs, most of the concepts seen in Native API tracing also apply, but there are also significant differences, both in the steps that were carried out and in the Pin APIs which were employed, as it can be observed from the diagram in \textit{Figure 3.5}.

@@

\textit{drltrace} (already mentioned in section 2.3) comes with its own external configuration file, in which each line contains information relative to an API, as it can be observed from the sample configuration file entry below (\textit{Listing 3.9}).      
\\
\lstset{
    language=C++,
    frame=single, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=none, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}
\begin{lstlisting}[caption={Sample entry in \textit{drltrace} configuration file},captionpos=b]
bool|GetUILanguageInfo|DWORD|wchar*|__out wchar*|__inout DWORD*|__out DWORD*
\end{lstlisting}
\lstset{
    language=C++,
    frame=tb, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=left, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}

@@

In order to extract information relative to the APIs from PyREbox's database, a Python script was employed, which, for every API, instantiates a \texttt{libcall\_info\_t struct} (\textit{Listing 3.10}) in a \texttt{.cpp} file.
\\
\begin{lstlisting}[caption={\texttt{struct} containing API-related information from PyREbox's database},captionpos=b]
typedef struct _libcall_info_t {
	const char* func_name;     			// API name
	int argnum;				   			// Number of Args
	libcall_arg_info_t lib_args[19];	// Array of arguments
} libcall_info_t
\end{lstlisting}
\vspace{1cm}

The most important field of such \texttt{struct} is surely the arguments' array, which is made up of \texttt{libcall\_arg\_info\_t} data structures (\textit{Listing 3.11}) containing the information relative each argument of an API, including an entry specific for the return value data. The array size was set to 19, since it was observed from the database data that the API with the most parameters had 18 arguments. This time, it was decided not to reuse NTAPI's \texttt{sysinfo\_arg\_t struct} (\textit{Listing 3.2}) to store the API's arguments information due to its lack of the field storing the argument name. 
\\
\begin{lstlisting}[caption={\texttt{struct} containing information associated to an API argument},captionpos=b]
typedef struct _libcall_arg_info_t {

	int arg_num;				// Arg Ordinal
	const char* arg_name;		// Arg Name
	int arg_type;             	// Arg Type
	char* arg_type_name;		// Arg Type Name
	int size;					// Size
	bool in_out_flag;			// Input/Output Flag

} libcall_arg_info_t;
\end{lstlisting}

@@

To give a better idea on how API data is represented, an instance of \texttt{libcall\_arg\_info\_t} containing the information associated to \texttt{WriteFileEx} can be observed below (\textit{Listing 3.12}).
\\
\begin{lstlisting}[caption={Instance of \texttt{libcall\_arg\_info\_t} relative to \texttt{WriteFileEx}},captionpos=b]
	{ "WriteFileEx",5,
	{ 
		{-1, "Return value", NKT_DBFUNDTYPE_SignedDoubleWord, 0, 4, INOUT },
		{0, "hFile", NKT_DBFUNDTYPE_UnsignedDoubleWord, 0, 4, IN },
		{1, "lpBuffer", NKT_DBFUNDTYPE_Void | NKT_DBOBJCLASS_Pointer, 0, 0, IN },
		{2, "nNumberOfBytesToWrite", NKT_DBFUNDTYPE_UnsignedDoubleWord, 0, 4, IN },
		{3, "lpOverlapped", NKT_DBOBJCLASS_Struct | NKT_DBOBJCLASS_Pointer,
		 "_OVERLAPPED", 160, INOUT },
		{4, "lpCompletionRoutine", NKT_DBOBJCLASS_Typedef, 0, 0, IN },
	}
	}
\end{lstlisting}

Due to the high amount of APIs for which information was stored in the \texttt{PyREbox} database, it was decided to group API data by DLL. Specifically, for every DLL a corresponding \texttt{.cpp} file was created to store the API-related information under the form of \texttt{libcall\_arg\_info\_t} instances, one for each "known" API exported by that DLL. Furthermore, in order to make API data retrieval more efficient, it was chosen to employ a two-level hash map approach for this purpose. Given the DLL name (easily obtainable in the image callback via the use of \texttt{IMG\_Name}), the first level hash map is utilized to retrieve the DLL-specific second level hash map, which, in turn, maps the API name to the appropriate \texttt{struct} instance (\textit{Figure 3.6}). This is how API information retrieval is carried out in case \textit{PyREbox} database is used as API data source.
\\
\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{Figures/HashMap.pdf}
\caption{\textit{PyREbox API Information Retrieval}}
\end{figure}

Let us conclude this section by explaining how analysis routines are inserted before and after the API execution during the API instrumentation procedure outlined in page 31. In BlueTracer such operation heavily relies on the use of the \texttt{RTN\_InsertCall}, which allows to insert analysis functions relative to a \texttt{RTN} object. This API takes as input the following parameters:
\begin{itemize}
\item The routine object (\texttt{RTN}) representing the function under analysis.
\item A member of the \texttt{IPOINT} enumeration which determines where the analysis function is inserted in relation to the instrumented object. It can be either \texttt{IPOINT\_BEFORE}, in which case the analysis code is placed before the instrumented function, or \texttt{IPOINT\_AFTER}, in which case the analysis code is placed after the instrumented function.
\item The analysis function to be called.
\item The arguments to pass to the analysis function, specified via members of the \texttt{IARG\_TYPE} enumeration. If the analysis function is placed before the instrumented object, then such arguments may also include the parameters' values of the function being instrumented. However, for this to occur the pair <\texttt{IARG\_FUNCARG\_ENTRYPOINT\_VALUE}, argument number> needs to be included for each of the analyzed function's parameters. Conversely, to obtain the instrumented function's result, \texttt{IARG\_FUNCRET\_EXITPOINT\_VALUE} is required. Obviously, this is possible only if the analysis function is placed after the instrumented object.
\end{itemize}

@@

Let us now dissect an invocation of \texttt{RTN\_InsertCall} to further clarify how BlueTracer adds analysis code before the execution of an API (\textit{Listing 3.13}).  
\begin{lstlisting}[caption={\texttt{RTN\_InsertCall} invocation to add analysis code before an API's execution},captionpos=b]
RTN_InsertCall(rtn, IPOINT_BEFORE, (AFUNPTR)PrintInfoBefore[Number of API Args],
 	IARG_FAST_ANALYSIS_CALL,
	IARG_ADDRINT, argNumber,
	IARG_ADDRINT, libPointer, IARG_ADDRINT, rtn_name, IARG_ADDRINT, img_name,
	IARG_REG_VALUE, REG_STACK_PTR,
	IARG_THREAD_ID,
	IARG_RETURN_IP,
	[Number of API Args]_ARGS
	IARG_END);
\end{lstlisting}

@@

} apicall_t;
\end{lstlisting}

The \texttt{apicall\_t struct} is initialized utilizing the API-related data, which was fetched beforehand during instrumentation and was passed to the analysis function via \texttt{libPointer}, as well as the values of the API arguments, which were provided to the analysis function through \texttt{IARG\_FUNCARG\_ENTRYPOINT\_VALUE} as already discussed in \textit{Section 3.4.1}.

@@

Therefore, in light of what just stated, the last operations which are done when tracing an API before its execution are allocating a shadow stack entry (\textit{Listing 3.15}), initializing it and placing it on top of the shadow stack in the thread local data, so that it can be retrieved after the API's execution.
\\
\begin{lstlisting}[caption={\texttt{struct} representing a shadow stack entry},captionpos=b]
typedef struct _stackEntry {
	apicall_t* apiInfo;    // API Call Reference
	ADDRINT  currentSP;    // Stack Pointer Register Content
	THREADID threadID;     // Thread ID
	uint api_counter;	   // Call Counter
} stackEntry;
\end{lstlisting}

@@

This API allows to register a notification function to execute just before the pinned application changes context due to the reception of an asynchronous event, like for instance a callback or a Windows APC. In BlueTracer, the idea is to use such registered notification function to extract and log information relative to the context change. The notification function in question is of type \texttt{CONTEXT\_CHANGE\_CALLBACK} and the most important arguments it takes as input are:
\begin{itemize}
\item A member of the \texttt{CONTEXT\_CHANGE\_REASON} enumeration indicating the reason for the context change. This is the parameter allowing BlueTracer to distinguish between callbacks and APCs. In fact, in the case of callbacks it is set to \texttt{CONTEXT\_CHANGE\_REASON\_CALLBACK}, while in the case of APCs it is set to \texttt{CONTEXT\_CHANGE\_REASON\_APC}. 
\item The application's register state before the context change. It is \texttt{null} if the context change is caused by a callback.
\item The application's register state after the context change.  
\end{itemize}

Having outlined the tools made available by Pin to deal with the application's context changes, let us now explain in detail how callbacks tracing and Windows APCs tracing is carried out in BlueTracer.

@@

In light of what just stated, it is natural for BlueTracer's logging activity, taking place in the registered \texttt{CONTEXT\_CALLBACK\_NOTIFICATION} function, to involve the record of a callback occurrence, together with the name of the callback that is about to be invoked. Unfortunately, tracing the callback's arguments would require further work since they are packed in a data structure, as previously explained.

Right before the execution of \texttt{KiUserCallbackDispatcher}, i.e. when the registered \texttt{CONTEXT\_CALLBACK\_NOTIFICATION} function is called, the stack looks as follows (\textit{Figure 3.7}).
\\
\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{Figures/StackCallback.pdf}
\caption{\textit{Stack before \texttt{KiUserCallbackDispatcher} execution}}
\end{figure}

@@

The resulting log format is exactly the same as the one adopted for user mode callbacks.

\iffalse
\section{Tool Configuration}
\fi

>>>>

In light of what just stated, BlueTracer's integration with BluePill is straightforward, since it essentially complements the \textit{Target platform simulation layer}, as it can be observed from \textit{Figure 3.1}.
\\
\begin{figure}[h]
\centering
\includegraphics[width=0.65\textwidth]{Figures/BluePill.pdf}
\caption{\textit{BlueTracer's integration with BluePill}}
\end{figure}

@@

In Pin, the Native API is intercepted right before \texttt{sysenter} is executed. This could be inferred by the fact that, in \texttt{TraceSysCallEntry}, the instruction pointer (obtainable via the \texttt{PIN\_GetContextReg} API) contains \texttt{sysenter}'s address. In light of this, in addition to the fact that each Native API calls \texttt{KiFastSystemCall} without setting a stack frame, the application's stack during the execution of \texttt{TraceSysCallEntry} is in the following state (\textit{Figure 3.4}).
\\
\begin{figure}[h]
\centering
\includegraphics[width=0.7\textwidth]{Figures/StackMainImage.pdf}
\caption{\textit{Stack before} \texttt{sysenter} \textit{execution}}
\end{figure}     
Having outlined the general situation, it is finally possible to describe how the main image check is carried out:

@@

		// Array of arguments
		sysinfo_arg_t arg[MAX_ARGS_IN_ENTRY];
		...
	} syscall_info_t;
\end{lstlisting}

@@

Even for \texttt{syscall\_tracer}, each field is self-explanatory. The only one which was not encountered before is the \texttt{syscall\_counter} field. The rationale behind this field is that, in the log, each traced call has a unique integer associated to it, utilized to group together the information relative to the call in post-processing. The next identifier to be assigned is stored in the \texttt{call\_number} field of the thread local storage (\textit{Listing 3.1}). Therefore, \texttt{syscall\_counter} simply contains the unique identifier associated to the Native API in the log. The arguments of a specific Native API are represented via the \texttt{drsys\_arg\_t} struct (\textit{Listing 3.6}).
\texttt{drsys\_arg\_t}'s fields are mostly the same of \texttt{sysinfo\_arg\_t}, with three additions:
\begin{itemize}
\item \texttt{pre}, which is a boolean flag set to \texttt{true} when the logging is occurring before the Native API is executed (i.e. in \texttt{TraceSysCallEntry} and set to false when the logging is occurring after the Native API is executed (i.e. in \texttt{TraceSysCallExit}).
\item \texttt{enum\_name}, a string containing the symbolic name of the enumeration a named constant belongs to. In most cases, this is the symbolic name associated to the first constant of the enumeration. Such field was introduced to decouple type name and enumeration name.
\item \texttt{value}, an \texttt{ADDRINT}\footnote{The ADDRINT type is defined by Pin and represents a memory address} containing the value of the argument. This is obtained through the \texttt{PIN\_GetSyscallArgument} API, which takes as input the context before the execution of the system call, the system call standard and the ordinal number of the argument whose value is requested.  
\end{itemize}
\begin{lstlisting}[caption={\texttt{struct} representing an argument of the Native API being traced},captionpos=b]
typedef struct _drsys_arg_t {

@@

@@

\begin{itemize}
\item Finally, if the argument's type is not a primitive one, it is checked if it is \texttt{null} and if it is a complex output parameter being traced before the Native API execution, situation in which only the address stored in the pointer is logged. If none of these two scenarios are true, then the tracer finds itself in the situation where it is required to log the value of a complex type. This is performed through the invocation of \texttt{print\_known\_compound}. As of right now, such function differentiates between four complex types, namely \texttt{UNICODE\_STRING}, \texttt{OBJECT\_ATTRIBUTES}, \texttt{IO\_STATUS\_BLOCK} and \texttt{LARGE\_INTEGER}, and logs the argument's value accordingly. If the argument's type is a complex type that is not supported, the symbolic value \texttt{<NYI>} (not yet implemented) is recorded.

@@

To conclude this section, let us show how a traced Native API looks like in the log file (\textit{Listing 3.8}). As it can be observed, the adopted log format is quite intuitive. In fact, the idea was to have a log which is both easy to read by the analyst but also easy to parse and post-process.      
\\

@@

In Microsoft Windows, APIs are implemented as functions provided by a set of dynamic-link libraries. These, also known as DLLs, constitute Microsoft's way of implementing the concept of shared libraries in the Windows operating system. Therefore, the functions they make available can be employed by different applications \cite{DLL}. Windows APIs offer a plethora of functionalities which belong to many different categories, ranging from base services (e.g. file management) to user interface functions and network operations \cite{WinAPI}. In light of this, a Windows API has rich semantic information associated to it and tracing the sequence of APIs a malware calls provides analysts with a very high-level view of the sample's behavior. This is why BlueTracer's API tracer component is undoubtedly the most important one. 
\begin{figure}[h]
\centering
\vspace{0.5cm}
\includegraphics[scale=0.662]{Figures/APITracing.pdf}
\caption{\textit{APIs Tracing Workflow}}
\end{figure}
When tracing APIs, most of the concepts seen in Native API tracing also apply, but there are also significant differences, both in the steps that were carried out and in the Pin APIs which were employed, as it can be observed from the diagram in \textit{Figure 3.5}.

@@

\textit{drltrace} (already mentioned in section 2.3) comes with its own external configuration file, in which each line contains information relative to an API, as it can be observed from the sample configuration file entry below (\textit{Listing 3.9}).      
\vspace{1cm}
\lstset{
    language=C++,
    frame=single, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=none, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}
\begin{lstlisting}[caption={Sample entry in \textit{drltrace} configuration file},captionpos=b]
bool|GetUILanguageInfo|DWORD|wchar*|__out wchar*|__inout DWORD*|__out DWORD*
\end{lstlisting}
\lstset{
    language=C++,
    frame=tb, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=left, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}

@@

In order to extract information relative to the APIs from PyREbox's database, a Python script was employed, which, for every API, instantiates a \texttt{libcall\_info\_t struct} (\textit{Listing 3.10}) in a \texttt{.cpp} file.
\\
\begin{lstlisting}[caption={\texttt{struct} containing API-related information from PyREbox's database},captionpos=b]
typedef struct _libcall_info_t {
	const char* func_name;     			// API name
	int argnum;				   			// Number of Args
	libcall_arg_info_t lib_args[19];	// Array of arguments
} libcall_info_t
\end{lstlisting}

The most important field of such \texttt{struct} is surely the arguments' array, which is made up of \texttt{libcall\_arg\_info\_t} data structures (\textit{Listing 3.11}) containing the information relative each argument of an API, including an entry specific for the return value data. The array size was set to 19, since it was observed from the database data that the API with the most parameters had 18 arguments. This time, it was decided not to reuse NTAPI's \texttt{sysinfo\_arg\_t struct} (\textit{Listing 3.2}) to store the API's arguments information due to its lack of the field storing the argument name. 
\\
\begin{lstlisting}[caption={\texttt{struct} containing information associated to an API argument},captionpos=b]
typedef struct _libcall_arg_info_t {
	int arg_num;				// Arg Ordinal
	const char* arg_name;		// Arg Name
	int arg_type;             	// Arg Type
	char* arg_type_name;		// Arg Type Name
	int size;					// Size
	bool in_out_flag;			// Input/Output Flag
} libcall_arg_info_t;
\end{lstlisting}

@@

\newpage
To give a better idea on how API data is represented, an instance of \texttt{libcall\_arg\_info\_t} containing the information associated to \texttt{WriteFileEx} can be observed below (\textit{Listing 3.12}).
\\
\begin{lstlisting}[caption={Instance of \texttt{libcall\_arg\_info\_t} relative to \texttt{WriteFileEx}},captionpos=b]
	{ "WriteFileEx",5,
	{ 
		{-1, "Return value", NKT_DBFUNDTYPE_SignedDoubleWord, 0, 4, INOUT },
		{0, "hFile", NKT_DBFUNDTYPE_UnsignedDoubleWord, 0, 4, IN },
		{1, "lpBuffer", NKT_DBFUNDTYPE_Void | NKT_DBOBJCLASS_Pointer, 0, 0, IN },
		{2, "nNumberOfBytesToWrite", NKT_DBFUNDTYPE_UnsignedDoubleWord, 0, 4, IN },
		{3, "lpOverlapped", NKT_DBOBJCLASS_Struct | NKT_DBOBJCLASS_Pointer,
		 "_OVERLAPPED", 160, INOUT },
		{4, "lpCompletionRoutine", NKT_DBOBJCLASS_Typedef, 0, 0, IN },
	}
	}
\end{lstlisting}

Due to the high amount of APIs for which information was stored in the \texttt{PyREbox} database, it was decided to group API data by DLL. Specifically, for every DLL a corresponding \texttt{.cpp} file was created to store the API-related information under the form of \texttt{libcall\_arg\_info\_t} instances, one for each "known" API exported by that DLL. Furthermore, in order to make API data retrieval more efficient, it was chosen to employ a two-level hash map approach for this purpose. Given the DLL name (easily obtainable in the image callback via the use of \texttt{IMG\_Name}), the first level hash map is utilized to retrieve the DLL-specific second level hash map, which, in turn, maps the API name to the appropriate \texttt{struct} instance (\textit{Figure 3.6}). This is how API information retrieval is carried out in case \textit{PyREbox} database is used as API data source.
\begin{figure}[h]
\centering
\vspace{0.5cm}
\includegraphics[width=1\textwidth]{Figures/HashMap.pdf}
\caption{\textit{PyREbox API Information Retrieval}}
\end{figure}

Let us conclude this section by explaining how analysis routines are inserted before and after the API execution during the API instrumentation procedure outlined in pages 34-35. In BlueTracer such operation heavily relies on the use of the \texttt{RTN\_InsertCall}, which allows to insert analysis functions relative to a \texttt{RTN} object. 
\newpage
This API takes as input the following parameters:
\begin{itemize}
\item The routine object (\texttt{RTN}) representing the function under analysis.
\item A member of the \texttt{IPOINT} enumeration which determines where the analysis function is inserted in relation to the instrumented object. It can be either \texttt{IPOINT\_BEFORE}, in which case the analysis code is placed before the instrumented function, or \texttt{IPOINT\_AFTER}, in which case the analysis code is placed after the instrumented function.
\item The analysis function to be called.
\item The arguments to pass to the analysis function, specified via members of the \texttt{IARG\_TYPE} enumeration. If the analysis function is placed before the instrumented object, then such arguments may also include the parameters' values of the function being instrumented. However, for this to occur the pair <\texttt{IARG\_FUNCARG\_ENTRYPOINT\_VALUE}, argument number> needs to be included for each of the analyzed function's parameters. Conversely, to obtain the instrumented function's result, \texttt{IARG\_FUNCRET\_EXITPOINT\_VALUE} is required. Obviously, this is possible only if the analysis function is placed after the instrumented object.
\end{itemize}

@@

Let us now dissect an invocation of \texttt{RTN\_InsertCall} to further clarify how BlueTracer adds analysis code before the execution of an API (\textit{Listing 3.13}).  
\\
\begin{lstlisting}[caption={\texttt{RTN\_InsertCall} invocation to add analysis code before an API's execution},captionpos=b]
RTN_InsertCall(rtn, IPOINT_BEFORE, (AFUNPTR)PrintInfoBefore[Number of API Args],
 	IARG_FAST_ANALYSIS_CALL,
	IARG_ADDRINT, argNumber,
	IARG_ADDRINT, libPointer, IARG_ADDRINT, rtn_name, IARG_ADDRINT, img_name,
	IARG_REG_VALUE, REG_STACK_PTR,
	IARG_THREAD_ID,
	IARG_RETURN_IP,
	[Number of API Args]_ARGS
	IARG_END);
\end{lstlisting}

@@

} apicall_t;
\end{lstlisting}
\vspace{1cm}
The \texttt{apicall\_t struct} is initialized utilizing the API-related data, which was fetched beforehand during instrumentation and was passed to the analysis function via \texttt{libPointer}, as well as the values of the API arguments, which were provided to the analysis function through \texttt{IARG\_FUNCARG\_ENTRYPOINT\_VALUE} as already discussed in \textit{Section 3.4.1}.

@@

Therefore, in light of what just stated, the last operations which are done when tracing an API before its execution are allocating a shadow stack entry (\textit{Listing 3.15}), initializing it and placing it on top of the shadow stack in the thread local data, so that it can be retrieved after the API's execution.
\vspace{3cm}
\begin{lstlisting}[caption={\texttt{struct} representing a shadow stack entry},captionpos=b]
typedef struct _stackEntry {
	apicall_t* apiInfo;    // API Call Reference
	ADDRINT  currentSP;    // Stack Pointer Register Content
	THREADID threadID;     // Thread ID
	uint api_counter;	   // Call Counter
} stackEntry;
\end{lstlisting}

@@

This API allows to register a notification function to execute just before the pinned application changes context due to the reception of an asynchronous event, like for instance a callback or a Windows APC. In BlueTracer, the idea is to use such registered notification function to extract and log information relative to the context change. The notification function in question is of type \texttt{CONTEXT\_CHANGE\_CALLBACK} and the most important arguments it takes as input are:
\begin{itemize}
\item A member of the \texttt{CONTEXT\_CHANGE\_REASON} enumeration indicating the reason for the context change. This is the parameter allowing BlueTracer to distinguish between callbacks and APCs. In fact, in the case of callbacks it is set to \texttt{CONTEXT\_CHANGE\_REASON\_CALLBACK}, while in the case of APCs it is set to \texttt{CONTEXT\_CHANGE\_REASON\_APC}. 
\item The application's register state before the context change. It is \texttt{null} if the context change is caused by a callback.
\item The application's register state after the context change.  
\end{itemize}      
Having outlined the tools made available by Pin to deal with the application's context changes, let us now explain in detail how callbacks tracing and Windows APCs tracing is carried out in BlueTracer.

@@

In light of what just stated, it is natural for BlueTracer's logging activity, taking place in the registered \texttt{CONTEXT\_CALLBACK\_NOTIFICATION} function, to involve the record of a callback occurrence, together with the name of the callback that is about to be invoked. Unfortunately, tracing the callback's arguments would require further work since they are packed in a data structure, as previously explained.
\newpage
Right before the execution of \texttt{KiUserCallbackDispatcher}, i.e. when the registered \texttt{CONTEXT\_CALLBACK\_NOTIFICATION} function is called, the stack looks as follows (\textit{Figure 3.7}).
\\
\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{Figures/StackCallback.pdf}
\caption{\textit{Stack before \texttt{KiUserCallbackDispatcher} execution}}
\end{figure}

@@

The resulting log format is exactly the same as the one adopted for user mode callbacks (\textit{Listing 3.17}).
\\
\lstset{
    language=C++,
    frame=single, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=none, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}
\begin{lstlisting}[caption={Log entry relative to a user mode APC},captionpos=b]
Context change caused by APC -> ifsutil.dll!?HardRead@IO_DP_DRIVE@@QAEEVBIG_INT@@KPAX@Z
\end{lstlisting}
\lstset{
    language=C++,
    frame=tb, % draw a frame at the top and bottom of the code block
    tabsize=4, % tab space width
    showstringspaces=false, % don't mark spaces in strings
    numbers=left, % display line numbers on the left
    commentstyle=\color{ao}, % comment color
    keywordstyle=\color{blue}, % keyword color
    stringstyle=\color{red}, % string color
    basicstyle=\footnotesize\ttfamily,
    basewidth = {.48em}
}
